---
title: "For Loops and If Else Statements"
author: "Dennis Teo"
date: "2020-06-13"
summary: "For loops and if else statements are core skills in R programming which allow you to solve problems flexibly. This tutorial covers the conceptual basics of how these functions work and demonstrates a few simple examples on how to use them."
tags: ["Tutorial", "R Core Skills"]
header-includes:
- \usepackage{setspace}
- \doublespacing
output:
  html_document:
    df_print: paged
    number_sections: yes
    theme: united
    highlight: tango
    toc: yes
    toc_depth: 3
    toc_float: true
fontsize: 16pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

<style> 
p {line-height:2em;} 
</style>

To advance your core skills in R (or any programming language), you need to have a good grasp of **for loops** and **if else statements**. These two functions do a lot of heavy lifting in many of the functions you used before. Once learned, you will definitely expand your ability to solve novel coding problems in R.  

Because of their utility, a quick Google search will yield you many existing tutorials on these functions. That said, I will include my own version of a tutorial for these functions. I have taught these functions before and people often pick up misconceptions on how these functions work. In this tutorial, I will attempt to slowly teach the mechanics of these functions and how they work. This tutorial is mainly aimed at people with little to no coding background where the slow pace of teaching is important.  

# for loops

A **for loop** is an iterative function. The basis is that it takes some code and repeats it a number of times based on your specification. Let's first develop an intuition of repeating code. Here's an example of counting sheep.  

```{r}

over.the.fence <- c() # creates an empty vector

over.the.fence <- c("sheep", over.the.fence) # add a sheep to the empty vector

print(paste(length(over.the.fence), "sheep over the fence")) # counting number of sheep

# Note:
# length() computes the number of elements in a vector
# paste() combines its inputs into a single character object
# print() prints out the result for us to view

```

As you can see, this code simply adds a sheep into the object, **over.the.fence**. What's neat about this code is that when I repeat it (through simple copy and paste), I increase my sheep count. Say I want to count 3 sheep, I can repeat this code 2 more times to get 3 sheep.  

```{r}

over.the.fence <- c("sheep", over.the.fence) # adding a sheep to the vector

print(paste(length(over.the.fence), "sheep over the fence")) # counting number of sheep

over.the.fence <- c("sheep", over.the.fence) # adding a sheep to the vector

print(paste(length(over.the.fence), "sheep over the fence")) # counting number of sheep

over.the.fence # our final object
```

Observe that I did not repeat the first line of code which **initializes** the empty object. If I repeated that, we will always have only one sheep since we keep emptying the vector before adding in a sheep.  

```{r}
over.the.fence <- c() # creating an empty vector

over.the.fence <- c("sheep", over.the.fence) # adding a sheep to the empty vector

print(paste(length(over.the.fence), "sheep over the fence")) # counting number of sheep

over.the.fence <- c() # creating an empty vector

over.the.fence <- c("sheep", over.the.fence) # adding a sheep to the empty vector

print(paste(length(over.the.fence), "sheep over the fence")) # counting number of sheep

over.the.fence # our final object

```

The above code essentially captures the logic of a **for loop**. We take a certain code (aka body of a function) and repeat it a number of times (in this case, 3 times). Often, we want to store our output into an object (over.the.fence). To do so, we need to initialize the object first (creating an empty vector). On each iteration, we modify this object (adding one sheep at a time) to eventually get the output that we want (3 sheep).    

Hopefully, you find this logic of iteration relatively straightforward. As we will see later and other examples throughout this tutorial, iteration can help us solve many problems. One issue of iteration, however, is that it is tedious. If I wanted to count 100 sheep using this method, I have to copy/paste my code 100 times. This is where a **for loop** becomes useful.  

Here is the template of how a **for loop** tends to look like. You begin with the **for()** function to begin the for loop. Next you specify the the number of times you wish to iterate the function using **(i in 1:N)** where N is the number of times you want to repeat the function. We will talk more about how this works later. Finally, within the {} brackets, you add the function you wish to iterate. Remember to close the bracket or your for loop will not run. (R helpfully does this for you but it is good to double-check as a habit).  

```{r eval = F, echo = T}

for (i in 1:N) {  # number of iterations
  
  # Body of the Function - you are repeating this
  
}


```

Now, let's use this template on the sheep counting problem.  

```{r}

over.the.fence <- c() # creating an empty vector

for (i in 1:10){ # repeat 10 times

over.the.fence <- c("sheep", over.the.fence) # adding a sheep to the vector

print(paste(length(over.the.fence), "sheep over the fence")) # counting number of sheep

}

over.the.fence # our final object

```

See that the pattern of the code follows the logic I described before. First, we initialize the object **outside** the **for loop**. This creates an object to store our herd of sheep. Then, we start the **for loop** to only repeat the code which adds sheep to our object. We replace N with 10 because we want to iterate our code 10 times to get 10 sheep over the fence.  

Initializing the object outside the for loop is important because if it is inside the for loop, it will be repeated as well. This is similar to the example I showed earlier where our sheep counts never increase because we keep emptying the vector we are adding sheep to.  

Okay. Let's get to the question some of you might be asking. What is **i**? Well, let's ask R what **i** is.

```{r}
i
```

See that **i** is 10, the number of times we specified the **for loop** to iterate. This is no coincidence. In this example, **i** acts as a counter for the loop. On each iteration, the value of **i** increased by 1.  

```{r}

for (i in 1:10){ # repeat 10 times
  
print(i) # print i on each iteration

}


```

By the way, the term **i** is not neccessary. It is common practice in coding to use terms like i, j, k, etc. However, this is not required for the code to run. 

```{r}

for (sheep.counter in 1:10){
  
print(sheep.counter) # print sheep.counter on each iteration

}


```

This property of the **for loop** is highly useful. One example is that in data cleaning, we often wish to repeat a function across the rows of our dataset. Using this counter allows us to match our function to the corresponding row of our datasets. Here is an example.  

```{r echo = F}

set.seed(100)

introversion.data <- data.frame(id = 1:10,
                 introversion1 = sample( 1:5, 10, replace = TRUE),
                 introversion2 = sample( 1:5, 10, replace = TRUE),
                 introversion3 = sample( 1:5, 10, replace = TRUE),
                 introversion4 = sample( 1:5, 10, replace = TRUE),
                 introversion5 = sample( 1:5, 10, replace = TRUE), stringsAsFactors = FALSE)

introversion.data
```

Say, we wish to compute a aggregate score of introversion (via summation) using the **for loop**.  

```{r}
nrow(introversion.data) # this gives us the number of rows in our dataset

introversion.items <- c("introversion1", "introversion2", "introversion3", "introversion4", "introversion5")

introversion.items # the names of our introversion items for aesthetic purposes

introversion.data$introversion.sum <- NA # initializing new column

for (i in 1:nrow(introversion.data)){ # for each row of the dataset
  
  introversion.data[i , "introversion.sum"] <- sum(introversion.data[i , introversion.items ]) # compute sum
  
}

introversion.data # showing the data

introversion.data$introversion.sum # showing the new computed variable only

```

Again, this follows the basic logic of iteration. First, we initialize an empty column which will house our computed scores later. It is good practice to initialize this column with NA.  

Next, we begin our **for loop**. The code we want to iterate is the summing of the introversion item scores for each participant. With the **i**, we can accomplish this. Observe that the **i** (counter for the loop) reflects each row of the dataset. On the first iteration, the **i** is 1 and it computes the sum for the first row of the dataset. On the second iteration, **i** is 2 and it computes the sum for the second row of the dataset. It does so until it reaches the total number of rows in the dataset (we use **nrow()** to specify this).  

Let's learn one last thing about the **for loop** before moving onto **if else statements**.  

When specifying the number of iterations, we tend to use **(i in 1:N)** as the default template. This is because it is intuitive and easy to keep track of the number of iterations using this method. We can even match the number of iterations to meaningful things like the number of rows in the dataset or the number of items in a vector (using the **length()** function).  

However, the **for loop** actually reads 1:N as a vector.  

```{r}
1:10
```

Instead, of increasing the value of **i** by one on each iteration, the **for loop** actually updates **i** to the next value in the numeric vector which you input.  

```{r}

for (i in 10:20 ){
  
print(i) # print i on each iteration

}


```

Here is another example which completely ignores sequencing.  

```{r}

for (i in c(10, 20, 5, 5, 7, 90) ){ 
  
print(i) # print i on each iteration

}


```

This property can actually help us on some problems. For example, let's say we only want to compute the sum of introversion scores for a subset of our participants. If we know the rows which correspond to these participants, we can re-write the **for loop** to only perform computations for this subset of participants.  

```{r}

compute.these.participants <- c(1, 3, 9, 10) 

compute.these.participants # vector which has the rows we want to compute

introversion.items # the names of our introversion items 

introversion.data$introversion.sum <- NA # initializing new column

for(i in compute.these.participants){ # iterate across only the rows that we want
  
  introversion.data[i , "introversion.sum"] <- sum(introversion.data[i , introversion.items ]) # compute sum
  
}

introversion.data # showing data set

introversion.data$introversion.sum # showing the new computed variable only

```

If we combine our knowledge of logical statements, we can further expand our capabilities with the for loop. Let's say we wish to only compute this aggregate score for participants who scored at least a 2 on introversion item 1. We can use **which()** to find out which rows correspond to these participants and run the **for loop** on these participants only.  

```{r}
which(c(TRUE, FALSE, TRUE, FALSE, TRUE)) # which() lets you find out the indexes of TRUE values 

which(introversion.data$introversion1 > 2) # this allows us to find the row numbers of participants with introversion1 score greater than 2

compute.these.participants <- which(introversion.data$introversion1 > 2) # let's save this into an object

introversion.items # the names of our introversion items for aesthetic purposes

introversion.data$introversion.sum <- NA # initializing new column

for(i in compute.these.participants){ # iterate across only the rows that we want
  
  introversion.data[i , "introversion.sum"] <- sum(introversion.data[i , introversion.items ]) # compute sum
  
}

introversion.data # showing data set

introversion.data$introversion.sum # showing the new computed variable only



```


# if else statements

**if else statements** allow you to specify the conditions you want to execute a set of functions. Here's an example. When evaluating p-values, we often apply a simple logic by convention. If p is less than .05, we reject the null hypothesis. Otherwise, we retain the null hypothesis. We can express this using an **if else statement**.  

```{r}

p <- .01 # assign p to the value .01 

if (p < .05) { # condition 1 - p less than .05
  
  print ("reject null hypothesis") # body 1 
  
  
} else { # all other cases
  
  print ("retain null hypothesis") # body 2
  
  
}

p <- .5 # assign p to the value .5

if (p < .05) { # condition 1 - p less than .05
  
  print ("reject null hypothesis") # body 1
  
  
} else { # all other cases
  
  print ("retain null hypothesis") # body 2
  
  
}

```

As you can see, depending on the value of p, the **if else statement** prints a different output. **if** (p < .05) specifies the condition to run the first body of code. If p < .05 evaluates to TRUE, the code runs body 1. If p < .05 evaluates to FALSE, then the code moves to the **else statement**. **else** represents *any other instance* when the first condition is not met.  

This is not the full extent of **if else statements**. In our example, the logic captures the basic convention of null hypothesis significance testing. However, it fails to handle some test cases appropriately.  

```{r}

p <- -1 # assign p to the value -1

if (p < .05) { # condition 1 - p less than .05
  
  print ("reject null hypothesis") # body 1
  
  
} else { # all other cases
  
  print ("retain null hypothesis") # body 2
  
  
}

```

Here, we have a case of an invalid input where the value of p is theoretically impossible (since p is a probability). However, the **if else statement** still returns an interpretable response because p = -1 is still less than .05.  

To handle invalid inputs, we need to add more conditions to our **if else statement**. To do so, we use the extension **else if**.  

```{r}

p <- -1 # assigns p to the value -1

if (p > 1 | p < 0){ # condition 1 - p more than 1 OR p less than 0
   
  print ("invalid input") # body 1

  } else if (p < .05) { # condition 2 - p less than .05
  
  print ("reject null hypothesis") # body 2
  
  
} else { # all other cases
  
  print ("retain null hypothesis") # body 3
  
  
}

```

To fix this issue, we first check if the input is more than 1 OR less than 0 since p should theoretically fall between 0 and 1. If the input meets this exclusion criteria, we note that the input is invalid.  

The sequence of these conditions are important. If the first condition is met (condition 1 returns TRUE), the **if else statement** will execute the first body of code. If not (condition 1 returns FALSE), the **if else statment** will check the second condition. The process repeats. If the second condition is met (condition 2 returns TRUE), the **if else statement** will execute the second body of code. Otherwise, it will move to the next step of the **if else statement**. As before, the **else** handles cases where all above conditions are not met.  

Following this logic, think about the scenario where we flip the order of conditions 1 and 2 for our current example.  

```{r}

p <- -1 # assign p to the value -1

if (p < .05){ # condition 1 - p less than .05
   
  print ("reject null hypothesis") # body 1

  } else if (p > 1 | p < 0) { # condition 2 - p more than 1 OR p less than 0
  
  print ("invalid input") # body 2
  
  
} else { # all other cases
  
  print ("retain null hypothesis") # body 3
  
  
}

```

This example illustrates the importance of ordering your conditions. Since p = -1 is also less than .05, the **if else statment** will execute the first body of function despite your attempt to exclude such inputs.  

Finally, the **else if** extension can be chained to handle more complex cases. In APA format, we have a specific convention for reporting p-values. If the value is smaller than .001, we typically report p < .001. In other cases, we report p-values in 2 decimal places. Let's try to put this logic into an **if else statement**.  

```{r}

p <- - 1 # assign p to the value -1

if (p > 1 | p < 0){ # condition 1 - p more than 1 OR p less than 0
   
  print ("invalid input") # body 1

  } else if (p < .001) { # condition 2 - p less than .001
  
  print ("p < .001") # body 2
    
  } else if ( p >= .001){ # condition 3 - p more than or equals to .001
    
    print (paste("p =", round(p , digits = 2))) # body 3 - round off to 2 decimal places
    
  } else { # all other cases
  
  NULL # body 4
  
}

```

```{r}

p <- .00002 # assign p to the value .00002

if (p > 1 | p < 0){ # condition 1 - p more than 1 OR p less than 0
   
  print ("invalid input") # body 1

  } else if (p < .001) { # condition 2 - p less than .001
  
  print ("p < .001") # body 2
    
  } else if ( p >= .001){ # condition 3 - p more than or equals to .001
    
    print (paste("p =", round(p , digits = 2))) # body 3 - round off to 2 decimal places
    
  } else { # all other cases
  
  NULL # body 4
  
  }

```

```{r}

p <- .0502 # assign p to the value .0502

if (p > 1 | p < 0){ # condition 1 - p more than 1 OR p less than 0
   
  print ("invalid input") # body 1

  } else if (p < .001) { # condition 2 - p less than .001
  
  print ("p < .001") # body 2
    
  } else if ( p >= .001){ # condition 3 - p more than or equals to .001
    
    print (paste("p =", round(p , digits = 2))) # body 3 - round off to 2 decimal places
    
  } else { # all other cases
  
  NULL # body 4
  
  }


```

As you might have noticed, we leave the body of the last condition (the else condition) blank by writing NULL. What makes **NULL** useful is that we can use it to skip certain conditions where we don't wish to perform any function.  

```{r}

p <- 1.5 

if (p > 1 | p < 0){
   
  NULL # this is skipper

  } else if (p < .001) { 
  
  print ("p < .001")
    
  } else if ( p >= .001){ 
    
    print (paste("p =", round(p , 2))) 
    
  } else {
    
    NULL
    
  }

```

One last tip about **if else statements** is that it is not necessary to specify the **else** when you have no need for it. For example, you can execute the following code without any problems.  

```{r}

count <- 0 # set count to 0

if (is.numeric(count)){ # if input is a number 
  
  count <- count + 1 # add count by 1
  
}

count

```

```{r}

count <- 10

if (is.numeric(count) & count > 0) { # if input is a number and greater than 0
  
  count <- count - 1 # subtract 1 from count
  
} else if (is.numeric(count) & count == 0) { # else if input is number and equals to 0
  
  print(paste("Countdown is Complete!")) # print message 
  
}

count

```

```{r}

count <- 0

if (is.numeric(count) & count > 0) { # if input is a number and greater than 0
  
  count <- count - 1 # subtract 1 from count
  
} else if (is.numeric(count) & count == 0) { # else if input is number and equals to 0
  
  print(paste("Countdown is Complete!")) # print message 
  
}

count

```

# combining for loops and if else statements

**for loops** and **if else statements** are both useful in their own right. However, they become especially powerful when used in conjunction. Many data cleaning problems can be creatively solved by using these together. Here, I will only give a short illustration of how this can be done. However, to really grasp the flexibility of these tools, you are highly encouraged to use them to solve your own problems.  

I will share two examples employing these 2 concepts. One involving vectors and one involving dataframes.  

## example involving vectors

For this example, we will go back to our sheep counting problem. This time, we have 3 different kinds of sheep, white sheep, black sheep, and bald sheep. We want to count the numbers of each type of sheep for any possible herd of sheep. Here's a randomly generated set of 100 sheep.  

```{r}

set.seed(100) # seed to ensure our "random" draws are fixed 

sheep.herd <- sample(c("white sheep", "black sheep", "bald sheep"), 100, replace = T) # randomly draw 100 sheep with replacement

sheep.herd # our herd

```

Now, we want to count the number of white sheep, black sheep, and bald sheep in this herd. To do so, we can use a **for loop** to iterate through this herd, sheep by sheep, keeping count of each type of sheep using an **if else statement**.    

```{r}

white.sheep.count <- 0 # initialize counter for white sheep
black.sheep.count <- 0  # initialize counter for black sheep
bald.sheep.count <- 0  # initialize counter for bald sheep

for (i in 1:length(sheep.herd)){ # for each sheep in a given herd
  
  if (sheep.herd[i] == "white sheep"){ # check if sheep is white
    
    white.sheep.count <-  white.sheep.count + 1 # add 1 to white sheep counter
    
  } else if (sheep.herd[i] == "black sheep") { # check if sheep is black
    
   black.sheep.count <-  black.sheep.count + 1 # add 1 to black sheep counter
    
  } else if (sheep.herd[i] == "bald sheep") { # check if sheep is bald
    
  bald.sheep.count <-  bald.sheep.count + 1 # add 1 to bald sheep counter
    
  } 
  
}
  
white.sheep.count # number of white sheep

black.sheep.count # number of black sheep

bald.sheep.count # number of bald sheep

```

First, we initialized a counter for each kind of sheep (starting at 0 sheep). Next, we iterated through the herd using a **for loop**. One by one, the **if else statement** checks for the kind of sheep. If it is a white sheep, we increase the count of white.sheep.count by 1. If it is a black sheep, we increase the count of black.sheep.count by 1 and so on.  

I omitted the **else** condition. Note that I could have written it such that there are only 2 conditions (white sheep and black sheep) and count the bald sheep after the **else** condition, like so.  

```{r}

white.sheep.count <- 0 # initialize counter for white sheep
black.sheep.count <- 0 # initialize counter for black sheep
bald.sheep.count <- 0 # initialize counter for bald sheep

for (i in 1:length(sheep.herd)){ # for each sheep in a given herd
  
  if (sheep.herd[i] == "white sheep"){ # check if sheep is white
    
    white.sheep.count <-  white.sheep.count + 1 # add 1 to white sheep counter
    
  } else if (sheep.herd[i] == "black sheep") { # check if sheep is black
    
   black.sheep.count <-  black.sheep.count + 1 # add 1 to black sheep counter
    
  } else { # count remaining sheep as bald
    
    bald.sheep.count <-  bald.sheep.count + 1 # add 1 to bald sheep counter
    
  }
  
}
  
white.sheep.count # number of white sheep

black.sheep.count # number of black sheep

bald.sheep.count # number of supposed bald sheep (all other entries are counted here as well)

```

While this yielded the correct counts, note that it is prone to error. If an input value wasn't "bald sheep", the code would still add a count to bald.sheep.count. This highlights the potential pitfall of the **else** statement. My recommendation is that you explicitly write out all conditions using ifs and else ifs as a default way of writing **if else statements**. Use the else statment to handle cases which you might have missed. I illustrate this point more clearly below.  

```{r}

set.seed(100)

sheep.herd <- sample(c("white sheep", "black sheep", "bald sheep", "horse"), 100, replace = T)

sheep.herd # new sheep herd but with horses

```

```{r}

white.sheep.count <- 0 
black.sheep.count <- 0
bald.sheep.count <- 0

others <- 0 # initialize counter for all other entries

for (i in 1:length(sheep.herd)){ 
  
  if (sheep.herd[i] == "white sheep"){
    
    white.sheep.count <-  white.sheep.count + 1 
    
  } else if (sheep.herd[i] == "black sheep") {
    
   black.sheep.count <-  black.sheep.count + 1
    
  } else if(sheep.herd[i] == "bald sheep") {
    
  bald.sheep.count <-  bald.sheep.count + 1
    
  } else { # count input that did not fall in any of the 3 kinds of sheep
    
    others <- others + 1
    
  }
  
}
  
white.sheep.count

black.sheep.count

bald.sheep.count

others # number of non-sheep entries

```


## example involving dataframes

Here, we will go back to our previous example of computing the aggregate score of introversion items. This time, however, I added an attention check column which represents whether participants carefully read the introversion items or not. (FALSE in cases where participants did not pay attention). Let's compute an aggregate score in cases where the attention check was met, leaving all other cases as NA. Of course, using **for loops** and **if else statements**.  

```{r echo = F}

introversion.items # the names of our introversion items for aesthetic purposes

introversion.data <- introversion.data[-7] # remove the aggregate column we computed earlier

introversion.data$attention.check <- sample(c(T, F), nrow(introversion.data), replace = T) # adding attention check

introversion.data

```

```{r}

introversion.data$introversion.sum <- NA # initialize output column

for (i in 1:nrow(introversion.data)){ # for each row in dataset
  
  if (introversion.data[i , "attention.check"] == TRUE){ # check if attention.check is TRUE
    
    introversion.data[i, "introversion.sum"] <- sum(introversion.data[i, introversion.items]) # compute sum
     
  } 
  
}

introversion.data # show dataset


```

We first initialize the output column (as NAs). Then, we iterate across the rows of the dataframe. On each row, we check if attention.check is TRUE. If it is, we compute the sum of the introversion items and store it in the output column (introversion.sum).  


## additional tips on using for loops and if else statements together

Since **for loops** and **if else statements** are often used together, there are some cool tricks that you can learn to use them together more effectively.  

As you know by now, **for loops** help you to iterate your code a set number of times. **if else statements** help to execute different codes based on the conditions you set. When used together, you are able to do a whole series of computations (via iteration), tweaking each computation according to the input (via if else) on each iteration of the **for loop**.  

However, under some conditions, you might wish to end the **for loop** immediately or skip some iterations of the **for loop**. This can be done using the **break** and **next** operator which saves computational time in some cases.  

### break 

Here I have a simple problem of finding out whether a name exists within a list of names.  

```{r}

list <- c("Bobby", "Tom", "Janet", "Tiffany", "Timmy", "Melissa")

list

```

Here, we have a short list of names. We might wish to find out whether a given name (e.g. Janet) is within this list. Of course, in this case, you can tell at a glance. However, if the list of names were longer, you might have more difficulty eyeballing the answer.  

Of course, this problem can be solved using many other methods (e.g. %in% operator, logical subsetting). However, you should think about how you might solve it using the tools we have learned in this tutorial.  

One simple strategy is to iterate through the list and return TRUE when a value in this list is equivalent to the input and FALSE otherwise.  

```{r}

input <- "Janet"

for (i in 1:length(list)){ # for each value in list
  
  if (list[i] == input) {
    
    print (TRUE)
    
  } else {
    
    print (FALSE)
    
  }
  
}

```

Now if you see TRUE appear in the output, you know that the input exists in the list.  

However, this code seems inefficient. All you really want to know is whether at least one of the items in the list matches the given input. It seems unnecessary to keep running the for loop after you have found a single match. In that event, you want to "break" out of the loop. To do so, we can use the **break** operator. 

```{r}

input <- "Janet"

for (i in 1:length(list)){ # for each value in list
  
  if (list[i] == input) {
    
    print (TRUE)
    
    break # break out of loop
    
  } else {
    
    print (FALSE)
    
  }
  
}

```

Notice that the output is slightly different. The loop forcibly ends when you add **break** into the **if else statement**.

### next 

Another similarly useful operator is the **next** operator. **next** allows you to "skip" to the next iteration of a **for loop** under conditions when you have no code you want to execute.  

Here is another simple example to illustrate the functionality of **next**. Let's go back to our sheep counting example.  

Say you have a sheep herd and wish to count the number of white sheep. You are not really concerned with the numbers of the other sheep. 


```{r}

set.seed(100)

sheep.herd <- sample(c("white sheep", "black sheep", "bald sheep"), 30, replace = T)

sheep.herd # new sheep herd but with horses

```

```{r}

white.sheep.count <- 0 

for (i in 1:length(sheep.herd)){ 
  
  if (sheep.herd[i] == "white sheep"){
    
    white.sheep.count <-  white.sheep.count + 1 
    
  } else { 
    
    next # skip to next iteration of loop
    
  }
  
  print(paste( white.sheep.count, "white sheep jumped over the fence" )) # print message
  
}

```

Notice that even though the print message is executed outside of the **if else** statement, we did not see any repeated sheep counts (see below for what happens without adding the **next** operator). This is because **next** forcibly ends the current iteration of the **for loop** and skips ahead to the next iteration. 


```{r}

white.sheep.count <- 0 

for (i in 1:length(sheep.herd)){ 
  
  if (sheep.herd[i] == "white sheep"){
    
    white.sheep.count <-  white.sheep.count + 1 
    
  } # did not add next
  
  print(paste( white.sheep.count, "white sheep jumped over the fence" )) # print message
  
}

```




# challenge yourself on an actual problem

We have come to the end of the tutorial on for loops and if else statements. If you wish to practice your skills on a challenging problem and see the use of these functions in action, see [here](https://msrcodelibrary.netlify.app/2020/06/13/compilation-of-for-loop-assignment-answers/).





