---
title: "Reading Qualtrics Survey"
author: "Daniel Foo"
date: "6/13/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here is a simple function to import Qualtrics CSV files into R. Here is a simple example of what it looks like when we upload a qualtrics CSV file: 

```{r echo=FALSE, message=FALSE, warning=FALSE}
#creating dummy dataset
qualtrics.file <- data.frame (ID = c("Participant ID", '{"ImportId":"QID164_TEXT"}', 1, 2, 3), 
                              Name = c("Participant Name", '{"ImportId":"QID164_TEXT"}', "Daniel", "James", "John"))

print (qualtrics.file)


```


As you can see, the two rows that come after the coloumn names are unnecessary and ideally we would want to remove them while reading in the data set. 

One might wonder, why not just (1) remove them from the original excel file beforehand or (2) upload the file first and then remove the two rows. The first step is certainly the most common and easiest way to do it. However, it is bad practice to manually delete anything from the original excel CSV folder as there can potentially be human error. The later suggestion can cause classification issues for your variables. In the above example, the column "ID" may be falsely classified as characters when it should in fact be numeric. 


Here is the raw code for the function: 
```{r echo=TRUE}
read_qualtrics <- function (path, skip) {
  require (tidyverse)  #Requires you to have the tidyverse package installed
  
  #Saving the column names into object df.names
  df.names <- read_csv (path) %>% 
    names ()
  
  #Loading in the dataset with read_csv() skiping n = skip 
  #Using df.names as the column names
  df <- read_csv (path, col_names = df.names, skip = skip)
  
  return (df)
}
```

path: the path of your file       
skip: number of rows you want to exclude (including the column names)


Example: 

```{r echo=TRUE, message=FALSE, warning=FALSE}
#saving the object shown above as a csv
write.csv (qualtrics.file, "qualtrics file.csv")

df <- read_qualtrics ("C:/Users/Msr Lab/Desktop/Locker/NRF/qualtrics file.csv", skip = 3)

print (df)

```
Note: Ignore column X1 as it is automatically generated by write.csv()



